On the Gravitational stability of gravito-turbulent accretion disks

### 2016-03-23 paper/model3d.tex
revising

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}\label{ideal_eos}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
{\bf For simplicity we refer to a single adiabatic index as that
  appears in the equation of state for both 2D and 3D disk models 
  \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
                             %for the same 
                             %disk in 2D 
%johnson: specify 3D gamma, then translate into 2D gamma because disk
%model is 2D 
}
The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

\subsubsection{Beta cooling}\label{beta_cool_model}
In our beta cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}.
\end{align}
Recall  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. 
{\bf Here, $T_\mathrm{irr}$}
%Here, 
%\begin{align}\label{tirr_def}
%  \tirr = \theta T_\mathrm{eq}
%\end{align}
is a reference temperature field. 
%parameterized as a 
%fraction $\theta$ of the equilibrium temperature field 
%$T_\mathrm{eq}(z)$. 
Physically, $\tirr$ {\bf may} the floor temperature set by, 
for example, stellar or background irradiation. {\bf

}

{\bf
Note that $\theta$  
should only be interpreted as an irradiation parameter when it is 
defined through Eq. \ref{tirr_def} and 
Eq. \ref{beta_cool} adopted as the cooling function. 
}

For beta cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align}\label{poly_vert} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(z=0)$ is the equilibrium mid-plane density. {\bf Note that
  $\Gamma$ is only relevant 
  to the 3D problem as it determines the disk vertical structure.} 
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align}

>>>>

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}\label{ideal_eos}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
{\bf For simplicity we only refer to the adiabatic index as that
  appears in the equation of state for both 2D and 3D disk models 
  \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
                             %for the same 
                             %disk in 2D 
%johnson: specify 3D gamma, then translate into 2D gamma because disk
%model is 2D 
}
The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

\subsubsection{Beta cooling}\label{beta_cool_model}
In our beta cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c},
\end{align}
{\bf where $T_\mathrm{irr}$ is a reference temperature field, and
recall}  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter.
%Here, 
%\begin{align}\label{tirr_def}
%  \tirr = \theta T_\mathrm{eq}
%\end{align}
%parameterized as a 
%fraction $\theta$ of the equilibrium temperature field 
%$T_\mathrm{eq}(z)$. 
Physically, $\tirr$ {\bf may} the floor temperature set by, 
for example, stellar or background irradiation.

For beta cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align}\label{poly_vert} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(z=0)$ is the equilibrium mid-plane density. {\bf Note that
  $\Gamma$ is only relevant 
  to the 3D problem as it determines the disk vertical structure.} 
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right),
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align} 
{\bf with 
  \begin{align}\label{tirr_def}
    \theta = \frac{\tirr}{T_\mathrm{eq}},
  \end{align}
   where $T_\mathrm{eq}(z)$ is the equilibrium temperature field.
   We shall consider vertically isothermal disks with $\Gamma=1$, as
  appropriate for the outer parts of irradiated protoplanetary disks
   \citep{chiang97}. In this case  
   $\alpha$, $T_\mathrm{eq}$ and $\theta$ are simply constants. 
    Note that $\theta$  
    should only be interpreted as an irradiation parameter when it is 
    defined through Eq. \ref{tirr_def}, in conjunction with adopting  
    Eq. \ref{beta_cool} as the cooling function. 
 }

### 2016-03-23 paper/paper3d.tex
revising

{\bf 
\subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
  In this work, we formally regard the gravito-turbulent
  disk described above as an equilibrium state, to which we apply    
  standard linear stability analysis. Then by `perturbations' we mean   
  deviations away from this gravito-turbulent 
  basic state\footnote{This should not be confused with 
    flucuations with respect to the 
    laminar disk that maintain the underlying
    gravito-turbulence.}. 
  We assume that instabilities defined this way  
  signifies disk fragmentation, since numerical simulations  
  show that the non-existance of steady gravito-turbulence imply    
  fragmentation.   
}

@@

%\acknowlegement
%S. Stahler 
%A. Youdin

>>>>

{\bf 
\subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
  In this work, we formally regard the gravito-turbulent
  disk described above as an equilibrium state, to which we apply    
  standard linear stability analysis. Thus by `perturbations' we mean   
  deviations away from this gravito-turbulent 
  basic state\footnote{This should not be confused with 
    flucuations with respect to the 
    laminar disk that maintain the underlying
    gravito-turbulence.}. 
  We assume instabilities defined this way  
  signifies disk fragmentation, since numerical simulations  
  show that the non-existance of steady gravito-turbulence imply    
  fragmentation.   
}

@@

%\acknowlegement
%S. Stahler 
%A. Youdin 
%paardekooper
%cathie clarke 
%lodato
%gammie
